Testing and Isolation Efficacy:\\Insights from a Simple Epidemic Model

### 2022-03-06 manuscript_BMB.tex
Responding to reviewers and incorporating comments

\newcommand{\fref}[1]{Fig.~\ref{#1}}
\newcommand{\appref}[1]{Appendix~\ref{app:#1}}
% \newcommand{\Rlogo}{R\xspace}
\newcommand{\percap}{\emph{per capita}\xspace}
\newcommand{\Rnum}{\ensuremath{\mathcal{R}_0}\xspace}
\newcommand{\covid}{COVID-19\xspace}
\newcommand{\pro}[1]{\ensuremath{\frac{\partial #1}{\partial \rho}}}
\newcommand{\pder}[2]{\ensuremath{\frac{\partial#1}{\partial#2}}} %\pder y{x}
\newcommand{\testing}[1]{\ensuremath{{\cal T}_{#1}}\xspace}
\newcommand{\testinghat}[1]{\ensuremath{\hat{\cal T}_{#1}}\xspace}
\newcommand{\testtarget}{\ensuremath{w\_{IS}}}
\newcommand{\oldtesttarget}{\ensuremath{\frac{w_I}{w_S}}\xspace}
\newcommand*\subtxt[1]{_{\textnormal{#1}}}
% \newcommand{\rev}[1]{\comment{red}{Revised}{#1}}
%% Roman subscript
\DeclareRobustCommand\_{\ifmmode\expandafter\subtxt\else\textunderscore\fi}

@@

In this paper, we have developed and analyzed a simple compartmental model that combines epidemiological dynamics---as defined by a simple SIR model---with the dynamics of testing and isolation. Our model is a caricature: it models the most basic feedbacks between epidemic and testing processes, but does not attempt to incorporate the many known complications of COVID-19 epidemiology (e.g., exposed, pre-symptomatic, and asymptomatic compartments \citep{kain2021chopping}; time-varying testing rates; behavioural dynamics \citep{weitz2020awareness}); Thus, it is most appropriate for assessing the \emph{qualitative} phenomena that arise from the interactions between transmission dynamics and testing, rather than for making quantitative predictions or guiding pandemic responses.

@@

@@

@@

Although testing and tracing is a key part of infection control strategies, mathematical epidemiologists have typically analyzed it with detailed models designed to inform particular public health efforts \citep{endo2020implication,hellewell2020feasibility,jenness2020modeling}, rather than analyzing simple but general models of the feedback between testing and transmission dynamics.
There have been several modeling studies of testing and tracing dynamics and their interaction with epidemiological dynamics. In the context of repeated screening and random testing of isolated populations (such as the members of a university), \cite{bergstrom2020frequency} provided analytical results quantifying the effects that proactive screening of asymptomatic individuals and isolation of confirmed-positive cases could have in reducing the spread of disease. 
\cite{rogers2021high} simulated a SEIR model with testing and isolation; they similarly suggest a strategy of rapid testing with antigen tests and the subsequent isolation of confirmed-positive individuals.
% reviewer1.com1
\cite{friston2021testing} model the effects of self-isolation on testing and tracing with a focus on projections under different testing and tracing scenarios. They conclude that the emergence of a second wave depends almost primarily on the rate at which immunity is lost and that it is necessary to track asymptomatic individuals in order to control the outbreak. 
Our modeling approach differs from these previous efforts in that it examines the effects of test-return rates and of different levels of testing focus, from random to highly targeted. We hope this paper will inspire further explorations of the fundamental properties of epidemic models that incorporate explicit testing processes.

>>>>

\newcommand{\fred}[1]{\comment{cyan}{Fred}{#1}}
\newcommand{\ali}[1]{\comment{magenta}{Ali}{#1}}
\newcommand{\jd}[1]{\comment{blue}{JD}{#1}}
\newcommand{\david}[1]{\comment{orange}{DJDE}{#1}}
\newcommand{\bmb}[1]{\comment{red}{BMB}{#1}}
\newcommand{\todo}[1]{\comment{red}{TODO}{#1}}
\newcommand{\new}[1]{\comment{red}{NEW}{#1}}
\newcommand{\needref}{\textcolor{red}{(REFS?)}}

\newcommand{\fref}[1]{Fig.~\ref{#1}}
\newcommand{\appref}[1]{Appendix~\ref{app:#1}}
% \newcommand{\Rlogo}{R\xspace}
\newcommand{\percap}{\emph{per capita}\xspace}
\newcommand{\Rnum}{\ensuremath{\mathcal{R}_0}\xspace}
\newcommand{\covid}{COVID-19\xspace}
\newcommand{\pro}[1]{\ensuremath{\frac{\partial #1}{\partial \rho}}}
\newcommand{\pder}[2]{\ensuremath{\frac{\partial#1}{\partial#2}}} %\pder y{x}
\newcommand{\testing}[1]{\ensuremath{{\cal T}_{#1}}\xspace}
\newcommand{\testinghat}[1]{\ensuremath{\hat{\cal T}_{#1}}\xspace}
\newcommand{\testtarget}{\ensuremath{w\_{IS}}}
\newcommand{\oldtesttarget}{\ensuremath{\frac{w_I}{w_S}}\xspace}
\newcommand*\subtxt[1]{_{\textnormal{#1}}}
\newcommand{\rev}[1]{\comment{red}{Revised}{#1}}
\renewcommand{\rev}[1]{{#1}}
%% Roman subscript
\DeclareRobustCommand\_{\ifmmode\expandafter\subtxt\else\textunderscore\fi}

@@

In this paper, we have developed and analyzed a simple compartmental model that combines epidemiological dynamics---as defined by a simple SIR model---with the dynamics of testing and isolation. Our model is a caricature: it models the most basic feedbacks between epidemic and testing processes, but does not attempt to incorporate the many known complications of COVID-19 epidemiology (e.g., exposed, pre-symptomatic, and asymptomatic compartments \citep{kain2021chopping}; time-varying testing rates; behavioural dynamics \citep{weitz2020awareness}); 
% reviewer2:com1
\rev{Further, our model ignores the fact that the testing and isolation parameters can change over time due to public health responses. In other words, our model makes an assumption of ``independence'': we assume that testing and isolation affect the epidemic progression, but we do not account for the fact that epidemic progression is also likely to affect testing and isolation strategies.}
Thus, it is most appropriate for assessing the \emph{qualitative} phenomena that arise from the interactions between transmission dynamics and testing, rather than for making quantitative predictions or guiding pandemic responses.

@@

This finding does, however, suggest that cautious behaviour by people waiting for test results might have appreciable effects on epidemic spread. This reinforces the idea that people waiting for test results should be urged to isolate and take other measures to prevent infecting other, or being infected. It also provides general support for the idea of institutional ``gateway'' testing which restricts individuals' contacts or access until one or more negative tests have been received \cite{Muller}.

@@

This theoretical phenomenon also highlights a practical point: keeping people waiting for tests that will come back negative can increase spread. The best solution is to increase test-processing speed, but it would also be worth exploring the option of encouraging people to test more than once, in case an early test was taken before an individual had enough virus to register as positive. Future work could profitably explore the question of when it is beneficial to encourage people to test more than once.

@@

Although testing and tracing is a key part of infection control strategies, mathematical epidemiologists have typically analyzed it with detailed models designed to inform particular public health efforts \citep{endo2020implication,hellewell2020feasibility,jenness2020modeling}, rather than analyzing simple but general models of the feedback between testing and transmission dynamics.
There have been several modeling studies of testing and tracing dynamics and their interaction with epidemiological dynamics. In the context of repeated screening and random testing of isolated populations (such as the members of a university), \cite{bergstrom2020frequency} provided analytical results quantifying the effects that proactive screening of asymptomatic individuals and isolation of confirmed-positive cases could have in reducing the spread of disease. 
\cite{rogers2021high} simulated a SEIR model with testing and isolation; they similarly suggest a strategy of rapid testing with antigen tests and the subsequent isolation of confirmed-positive individuals.
\cite{friston2021testing} model the effects of self-isolation on testing and tracing with a focus on projections under different testing and tracing scenarios. They conclude that the emergence of a second wave depends primarily on the rate at which immunity is lost and that it is necessary to track asymptomatic individuals in order to control the outbreak. 
Our modeling approach differs from these previous efforts in that it examines the effects of test-return rates and of different levels of testing focus, from random to highly targeted. We hope this paper will inspire further explorations of the fundamental properties of epidemic models that incorporate explicit testing processes.

%% \fred{ our model makes the counterintuitive suggestion that increased testing rate could increase $\Rnum$. However, note that in our model, we made the assumption that once an individual is in the ``waiting'' compartment, they cannot be tested again until the test result returns. Thus, when the test return time is very long, this creates a massive time-delay where an individual may become infected while waiting for the test result, and then the test result comes back as negative, and so they believe they are negative when they are actually positive (and so they do not self-isolate at all, under the belief that they are negative).
%% So, in this case, it would be beneficial to actually re-test people even if their first test result has not come back. In fact, one could probably conceive of ``testing breadth'' (which roughly corresponds to the number of different people you test over some period of time) vs. ``testing depth (which corresponds to the number of times you test a single person over a period of time). Note that testing breadth gives us granular information ``in space but not in time'' (i.e., it tells us how many people are or are not infected at a single point in time), whereas testing depth gives us granular information ``in time but not in space''. The pathological situation in our model can be viewed as an imbalance between testing depth and breadth, where the testing strategy is overly-broad but not sufficiently deep.
%% In particular, one could build models that differentiate/partition ``testing intensity'' into ``testing breadth'' and ``testing depth''. In such a model, there should always exist some allocation of breadth and depth that makes it so that increasing the overall testing intensity is beneficial (i.e., if we choose the right strategy, more testing should always be good).
%% The most important point of all this: while the exact effect we identified in our model (that more testing can increase $\Rnum$ if done incorrectly) is unlikely to happen in real life, a weaker ``sister effect'' probably happens in real life all the time: we may be allocating our testing resources in the wrong breadth vs. depth ratio. Our model ``proves'' that pure-breadth is not always the optimal strategy (because, in fact, it can increase $\Rnum$). So, a more realistic future question is: what is the optimal allocation between breadth and depth that decreases $\Rnum$ most sharply for a given overall testing intensity? For example, in our model, when we increase testing intensity, we are mostly increasing breadth. So, even on parameter sets in our model where increasing rho decreases $\Rnum$, we may still be ``missing out'' by focusing only on breadth and not depth. }
%% \david{These are thoughtful ideas from Fred.  Perhaps they could be compressed to a couple of sentences describing a future direction worth pursuing.}

### 2022-03-06 response1_BMB.tex
Responding to reviewers and incorporating comments

\res 1 
We now have added several lines and the suggested reference [1] to the end of the discussion to acknowledge the ``gateway testing'' and the potential benefit of our modelling results in controling the disease spread.    
% {search for reviewer1:com1 in the manuscript_BMB.tex}

@@

[1] K. Muller and P. A. Muller, ``Mathematical modelling of the spread of covid-19 on a university campus,'' Infectious Disease Modelling, vol. 6, pp. 1025-1045, 2021.

>>>>

\res 1 
% {see "Muller" in manuscript_BMB.tex}

@@

% [1] K. Muller and P. A. Muller, ``Mathematical modelling of the spread of covid-19 on a university campus,'' Infectious Disease Modelling, vol. 6, pp. 1025-1045, 2021.
% ~/Downloads/MullMull.pdf
% https://pubmed.ncbi.nlm.nih.gov/34414342/
